---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# OSBNG

An R library for Ordnance Survey's British National Grid (BNG) index system.
This library provides tools for working with the BNG, a rectangular Cartesian
grid system used to identify and index locations across Great Britain into grid
squares at various resolutions.

## Overview

The osbng R package provides a programmatic interface to the British National
Grid (BNG), supporting efficient grid-based indexing and spatial analysis. This
enables applications such as statistical aggregation, data visualisation, and
data interoperability using BNG grid references. Designed for developers working
with geospatial data in the context of Great Britain, the library offers tools
to simplify working with the BNG, facilitating both technical integration into
geospatial workflows and practical exploration of the index system's structure.

The package supports the 'standard' BNG metre-based resolutions, which represent
powers of ten from 1m to 100km (1m, 10m, 100m, 1km, 10km, 100km). It also
supports the 'intermediate' quadtree resolutions (5m, 50m, 500m, 5km, 50km),
identified by an ordinal (NE, SE, SW, NW) BNG reference direction suffix.


## Installation

Install `osbng` in R directly from CRAN:

```{r, eval=FALSE}
install.packages("osbng")
```

Alternatively, you can install the latest version from
[GitHub](https://github.com/OrdnanceSurvey/osbng-r) with:

```{r, eval=FALSE}
remotes::install_github("OrdnanceSurvey/osbng-r")
```

Once installed, load the package to use it:

```{r, eval=FALSE}
library(osbng)
```


### Complimetary tools

* See the Python version: [`osbng-py`](https://github.com/OrdnanceSurvey/osbng-py)
* BNG grid data in GeoParquet and GeoPackage formats: [`osbng-grids`](https://github.com/OrdnanceSurvey/osbng-grids)

## Usage

The `osbng` package supports different interactions with the BNG index system
(e.g. indexing, hierarchy, traversal). A high-level summary of each task is
provided below:

### BNG Reference

`osbng` implements a custom BNG reference object, `BNGReference`. The
`BNGReference` object validates the reference and provides access to functions
to work with the reference.

```{r}
library(osbng)

bng_ref <- as_bng_reference("ST57SE")

print(bng_ref)

get_bng_resolution(bng_ref)

get_bng_resolution_string(bng_ref)
```

### Indexing

Provides the ability to index and work with coordinates and geometries against the BNG index system. This includes:

* Encoding easting and northing coordinates into `BNGReference` objects at a specified resolution.
* Decoding `BNGReference` objects back into coordinates, bounding boxes and grid squares as [`geos`](https://cran.r-project.org/package=geos) geometries.
* Indexing bounding boxes and `geos` geometries into grid squares at a specified resolution for spatial analysis.

The following example demonstrates a round trip of constructing a `BNGReference`
object from easting and northing coordinates, and then decoding back into
coordinates and a bounding box:

```{r}
bng_ref <- xy_to_bng(easting = 356976, northing = 171421, resolution = "5km")

bng_to_xy(bng_ref, position = "lower-left")

bng_to_bbox(bng_ref)
```

### Hierarchy

Provides functionality to navigate the hierarchical structure of the BNG index
system. This includes:

* Returning parents and children of `BNGReference` objects at specified resolutions.

The following example demonstrates the construction of a `BNGReference` object
from a reference string and returns a parent of the `BNGReference`:

```{r}
bng_ref <- as_bng_reference("ST5671SE")

bng_to_parent(bng_ref, resolution = "10km")
```

## Contributing

Please raise an issue to discuss features, bugs, or general questions.

## License

`osbng-r` is licensed under the terms of the MIT License.

