---
title: "osbng Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{osbng Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE
)
```

The `osbng` R package support geospatial grid indexing and interaction with
Ordnance Survey's British National Grid (BNG) index system. `osbng` uses
[`geos`](https://paleolimbot.github.io/geos/) for geospatial operations, with
[`sf`](https://r-spatial.github.io/sf/) an optional dependency. The `osbng`
package provides functionality supporting different interactions with the BNG
index system (e.g. indexing, hierarchy, traversal). This notebook provides an
overview of the package functionality, including example usage.

```{r setup}
library(osbng)
library(sf)
```

## BNG Index System

The OS BNG index system, also known as the OS National Grid, is a rectangular
Cartesian 700 x 1300km grid system based upon the transverse Mercator
projection. In the BNG, locations are specified using coordinates, eastings (x)
and northings (y), measured in meters from a defined origin point (0, 0)
southwest of the Isles of Scilly off the coast of Cornwall, England. Values
increase to the northeast, covering all of mainland GB and surrounding islands.

The BNG is structured using a hierarchical system of grid squares at various
resolutions. At its highest level, the grid divides GB into 100 km by 100 km
squares, each identified bya two-letter code. Successive levels of resolution
further subdivide the grid squares into finder detail, down to individual
1-meter squares.

```{r}
bng_bounds
```

```{r}
bng_prefixes
```

## Resolutions

`osbng` supports the "standard" BNG meter-based resolutions, which represent
powers of ten from 1m to 100km `(1m, 10m, 100m, 1km, 10km, 100km)`. It also
supports the "intermediate" quadtree resolutions `(5m, 50m, 500m, 5km, 50km)`,
identified by an ordinal `(NE, SE, WS, NW)` BNG reference direction suffix. The
`resolution` argument for functions can be expressed as a meter-based integer or
as a label.

```{r}
list_bng_resolution(lbl = TRUE)
```

```{r}
list_bng_resolution(lbl = FALSE)
```

## `BNGReference`

The BNG index system uses BNG references, also known more simply as grid or tile
references, to identify and index locations across Great Britain (GB) into grid
squares at various resolutions. The `osbng` package implements a custom
`BNGReference` object. This object validates and encapsulates a BNG reference,
providing functions to access and manipulate the reference. All functions
accepting or returning BNG references enforce the use of this class. Functions
within `osbng` are vectorised and compatible with pipes (`%>%`, `|>`) for
efficient processing of multiple BNG references.

```{r}
# Convert a BNG reference string into a `BNGReference` object
bng_ref <- as_bng_reference("SU 372 155")

bng_ref
```

```{r}
# The BNG reference string as a compact string
print(bng_ref, compact = TRUE)
```

```{r}
# The BNG reference string as a formatted string (default behaviour)
print(bng_ref, compact = FALSE)
```

```{r}
# The BNG resolution expressed in metres
get_bng_resolution(bng_ref)
```

```{r}
# The BNG resolution expressed as a descriptive label
get_bng_resolution_string(bng_ref)
```

## Indexing

A core component of `osbng` is the ability to index and work with coordinates
and geometries against the BNG index system. This includes:

* Encoding easting and northing coordinates into `BNGReference` objects at a specified resolution.
* Decoding `BNGReference` objects back into coordiantes, bounding boxes, and grid squares as `geos` geometries.
* Indexing boudning boxes and `geos` geometries into grid squares at a specified resolution for spatial analysis.

```{r}
# Easting and northing coordinates
x <- 437293
y <- 115543

# Convert easting and northing to BNGReference at 1km resolution
bng_ref <- xy_to_bng(x, y, resolution = "1km")

bng_ref
```

```{r}
# Decode BNGReference back into coordinates
# Default "position" is "lower-left"
bng_to_xy(bng_ref)
```

```{r}
# Decode BNGRefernece back into coordinates
# Centroid of the grid square
# "position" can be one of "lower-left", "upper-left", "lower-right", "upper-right", "centre"
bng_to_xy(bng_ref, position = "centre")
```

```{r}
# Return the grid square bounding box coordinates for the BNGRference object
bng_to_bbox(bng_ref)
```

```{r}
# Return the grid square as a geos geometry object for the BNGReference object
# Alternatively return the geometry as WKT or `sf` object
bng_to_grid_geom(bng_ref)
```

### Indexing `geos` Geometries

The `geom_to_bng` and `geom_to_bng_intersection` functions enable the indexing
of geometries, represented using `geos` Geometry objects, into grid squares at a
specified resolution. Both functions accept `geos` Geometry objects of the
following types: `Point`, `LineString`, `Polygon`, `MultiPoint`,
`MultiLineString`, `MultiPolygon`, and `GeometryCollection`. The coodinates must
be encoded in the [British National Grid (OSGB36)
EPSG:27700](https://epsg.io/27700) coordinate reference system.

These functions facilitate grid-based spatial analysis, enabling applications such as statistical aggregation, data visualisation, and data interoperability. The two functions differ in their operation:

* `geom_to_bng` returns the BNG grid squares intersected by the input geometry.
* `geom_to_bng_intersection` returns the intersections (shared geometries) between the input geometry and the grid square geometries.

See the vignette indexing for more examples and etail on the operation of the
two functions.

#### Point

```{r}
# Easting and norhting coordinates
x <- 437293
y <- 115543

# Construct a geos Point geometry from easting and northing coordinates
geom <- geos_make_point(x, y)

# Return the BNGReference object for the geometry at a 5km resolution
geom_to_bng(geom = geom, resolution = "5km")
```

```{r}
# Return the indexed results for the geometry at a 5km resolution
# For a Point geometry, the intersection geometry is the same as the original geometry
# The is_core property will always be False for a Point geometry
geom_to_bng_intersection(geom = geom, resolution = "5km")
```

#### LineString

```{r}
# Create a geos LineString geometry from Well Known Text (WKT)
geom <- geos_read_wkt("LINESTRING (436171.8 114893.7, 437861.3 116130.0)")

# Return a list of the BNGReference object(s) intersected by the geometry at a 1km resolution
geom_to_bng(geom = geom, resolution = "1km")
```

```{r}
# Return a nested list of the indexing results for the geometry at a 1km resolution
# The is_core property will always be False for a LineString geometry
geom_to_bng_intersection(geom = geom, resolution = "1km")
```

##### Visualise the results

For this example of visualising the results, we will use the `sf` library for
spatial data handling, which is a suggested package that works well with
`osbng`.

```{r, tidy=FALSE}
# Return the indexed results
bng_idx_geoms <- geom_to_bng_intersection(geom = geom, 
                                          resolution = "1km", 
                                          format = "sf")

# Store the indexed results as a spatial data frame
bng_idx_geoms <- st_sf(data.frame(bng_idx_geoms))

# Store the intersected BNGReference grid squares for context
bng_grid_geoms <- bng_to_grid_geom(bng_idx_geoms$bng_reference, 
                                   format = "sf")
```

```{r, tidy=FALSE, fig.height=6.5, fig.width=6.5}
# Categorical colours
# https://github.com/OrdnanceSurvey/GeoDataViz-Toolkit/blob/master/Colours/GDV%20colour%20palettes_v0.9.1.pdf
catcols <- c("#FF1F5B", "#009ADE", "#FFC61E", "#AF58BA")

# Plot geometries representing the intersection between the geometry and the BNG grid squares
plot(st_geometry(bng_idx_geoms),
     col = catcols,
     lwd = 3,
     extent = st_bbox(bng_grid_geoms),
     main = "Decomposition of LineString Geometry into BNG Grid Squares at a 1km Resolution",
     cex.main = .8,
     axes = TRUE,
     xlab = "Easting",
     ylab = "Northing")

# Plot intersected BNGReference grid squares for context
plot(st_geometry(bng_grid_geoms),
     col = NA,
     border = "#00000075",
     lty = "dashed",
     add = TRUE)
```

#### Polygon

```{r}
# Create a geos Polygon geometry from Well Known Text (WKT)
geom <- geos_read_wkt("POLYGON ((436661.45305455 115784.01571607, 437629.10896848 116112.11767069, 438229.486575 115953.45344189, 437990.51082297 114935.84767816, 436630.11197232 115085.69722758, 436661.45305455 115784.01571607))")
```

```{r}
# Return a list of the BNGRefernce objects intersected by the geometry at a 500m resolution
geom_to_bng(geom, resolution = 500)
```

```{r}
# Return a nested list of the indexed geometry and BNGReference objects for the geometry at 500m resolution
geom_to_bng_intersection(geom, resolution = 500)
```

##### Visualise the results

```{r}
# Return the indexed results
bng_idx_geoms <- geom_to_bng_intersection(geom = geom, 
                                          resolution = 500, 
                                          format = "sf")

# Store the indexed results as a spatial data frame
bng_idx_geoms <- st_sf(data.frame(bng_idx_geoms))

# Store the intersected BNGReference grid squares for context
bng_grid_geoms <- bng_to_grid_geom(bng_idx_geoms$bng_reference, 
                                   format = "sf")
```

```{r, tidy=FALSE, fig.height=6.5, fig.width=6.5}
# Plot core and edge geoemtries
plot(bng_idx_geoms["is_core"],
     border = "#fff",
     col = c("#009ade", "#ff1f5b")[bng_idx_geoms$is_core + 1],
     main = "Decomposition of Polygon Geometry into BNG Grid Squares at a 500m Resolution",
     cex.main = .8,
     axes = TRUE,
     xlab = "Easting",
     ylab = "Northing")

legend(0.05, 1, 
       title = "is_core",
       legend = c("True", "False"),  
       fill = c("#ff1f5b", "#009ade"))
```

## Hierarchy

`osbng` provides functionality to navigate the hierarchical structure of the BNG
index system. Traversal of this hierarchy is enabled by functions providing
methods to return the parent and children of `BNGReference` objects at specified
resolutions.

Parent and child definitions:

* **Parent**: The parent of a `BNGReference` object is the grid square at the next higher (i.e. coarser) resolution level that contains the current reference. For example, the parent of a 1km grid square reference would be the 5km grid square that contains it.

* **Children**: The children of a `BNGReference` object are the grid squares at the next lower (i.e. finer) resolution level that are contained within the current reference. For example, the children of a 10km grid square reference would be the 5km grid squares that it contains.

* While parent and child derivation defaults to the next higher and lower lower resolution, respectively, **any** supported resolution in the hiearchy can be specified.

```{r}
# Convert a BNG reference string into a BNGReference object
bng_ref <- as_bng_reference("SU 372 155")

# The resolution of the BNGReference object as a descriptive label
get_bng_resolution_string(bng_ref)
```

### Parent

```{r}
# Return the parent BNGReference object at the next higher resolution
bng_to_parent(bng_ref)
```

```{r}
# Return the parent BNGReference object at a custom higher resolution
bng_to_parent(bng_ref, resolution = "50km")
```

#### Visualise the results

```{r, tidy=FALSE, fig.height=6.5, fig.width=6.5}
# Return the parent BNGReference object
bng_ref_parent <- bng_to_parent(bng_ref)

# Store the BNGReference grid squares
bng_geom <- bng_to_grid_geom(bng_ref, format = "sf")
bng_parent_geom <- bng_to_grid_geom(bng_ref_parent, format = "sf")

# Plot the parent BNGReference grid square
plot(st_geometry(bng_parent_geom),
     col = "gray80",
     border = "#00000075",
     lty = "dashed",
     lwd = 2,
     main = "500m Resolution Parent of a 100m Resolution BNG Grid Square",
     cex.main = .9,
     axes = TRUE,
     xlab = "Easting",
     ylab = "Northing")

# Plot the original BNGReference grid square
plot(st_geometry(bng_geom),
     col = "#AF58BA",
     add = TRUE)
```

### Children

```{r}
# Return the children of the BNGReference object at the next lower resolution
bng_to_children(bng_ref)
```

```{r}
# Return the children of the BNGReference object at a custom lower resolution
bng_to_children(bng_ref, resolution = "10m")
```

#### Visualise the results

```{r, tidy=FALSE, fig.height=6.5, fig.width=6.5}
# Return the BNGReference object for the children at 10m resolution
bng_refs_children <- bng_to_children(bng_ref, resolution = "10m")

# Store the BNGReference grid squares
bng_geom <- bng_to_grid_geom(bng_ref, format = "sf")
bng_children_geom <- bng_to_grid_geom(bng_refs_children[[1]], format = "sf")

# Plot the parent BNGReference grid square
plot(st_geometry(bng_children_geom),
     col = "gray80",
     border = "#00000075",
     lty = "dashed",
     main = "10m Resolution Children of a 100m Resolution BNG Grid Square",
     cex.main = .9,
     axes = TRUE,
     xlab = "Easting",
     ylab = "Northing")

# Plot the original BNGReference grid square
plot(st_geometry(bng_geom),
     col = NA,
     border = "#AF58BA",
     lwd = 3,
     add = TRUE)
```

## Traversal

Traversal functionality in `osbng` includes calculating distances and neighbours
within the BNG index system. It supports spatial analyses such as
distance-constrained nearest neighbour searches and "distance within" queries by
offering:

* Generation of k-discs and k-rings around a give grid square.
* Identification of neighbouring grid squares and checking adjacency.
* Calculating the distance between grid square centroids or edges.
* Retrieving all grid squares within a specified absolute distance.

```{r}
# Convert a BNG reference string into a BNGReference object
bng_ref <- as_bng_reference("SU 372 155")
```

### K-disc

```{r}
# Return a list of BNGReference objects representing a filled disc around the BNGReference object
#  up to a grid distance k, including the given central BNGReference object
bng_kdisc(bng_ref, k = 2)
```

```{r, tidy=FALSE, fig.height=6.5, fig.width=6.5}
# Return a k-disc of the BNGReference object at a grid distance of k=2
bng_ref_kdisc = bng_kdisc(bng_ref, k = 2)

# Store the BNGReference grid squares
bng_geom <- bng_to_grid_geom(bng_ref, format = "sf")
bng_kdisc_geoms <- bng_to_grid_geom(bng_ref_kdisc[[1]], format = "sf")

# Plot the BNGReference grid squares in the k-disc
plot(st_geometry(bng_kdisc_geoms),
     col = "gray80",
     border = "#00000075",
     lty = "dashed",
     main = "K-Disc Around a BNG Grid Square Where k=2",
     cex.main = .9,
     axes = TRUE,
     xlab = "Easting",
     ylab = "Northing")

# Plot the original BNGReference grid square for context
plot(st_geometry(bng_geom),
     col = "#009ade",
     add = TRUE)
```

### K-ring

```{r}
# Return a list of BNGReference objects representing a hollow ring around around 
#  the BNGReference object at a grid distance k
bng_kring(bng_ref, k = 1)
```

```{r, tidy=FALSE, fig.height=6.5, fig.width=6.5}
# Return a k-ring of the BNGReference object at a grid distance of k=2
bng_ref_kring = bng_kring(bng_ref, k = 3)

# Store the BNGReference grid squares
bng_geom <- bng_to_grid_geom(bng_ref, format = "sf")
bng_kring_geoms <- bng_to_grid_geom(bng_ref_kring[[1]], format = "sf")

# Plot the BNGReference grid squares in the k-disc
plot(st_geometry(bng_kring_geoms),
     col = "gray80",
     border = "#00000075",
     lty = "dashed",
     main = "K-Ring Around a BNG Grid Square Where k=3",
     cex.main = .9,
     axes = TRUE,
     xlab = "Easting",
     ylab = "Northing")

# Plot the original BNGReference grid square for context
plot(st_geometry(bng_geom),
     col = "#009ade",
     add = TRUE)
```

### Neighbours

```{r}
# Convert BNG reference strings into BNGReference objects
bng_ref1 <- as_bng_reference("SU 372 155")
bng_ref2 <- as_bng_reference("SU 371 155")
```

```{r}
# Check if the two BNGReference objects are neighbours
bng_is_neighbour(bng_ref1, bng_ref2)
```

### Distance

```{r}
# Convert BNG reference strings into BNGReference objects
bng_ref1 <- as_bng_reference("SX")
bng_ref2 <- as_bng_reference("SU 1 2")

# Return the distance between two BNGReference objects in metres
# The default distance is calculated as the Euclidean distance between the centroids of the grid squares
bng_distance(bng_ref1, bng_ref2)
```

```{r}
# Return the distance between two BNGReference objects in metres
# edge_to_edge=True calculates the distance between the edges of the grid squares
bng_distance(bng_ref1, bng_ref2, edge_to_edge = TRUE)
```

```{r}
# Convert BNG reference stringsinto a BNGReference object
bng_ref <- as_bng_reference("ST 3 8")

# Return all BNGReference objects within a given distance d of the input BNGReference object
bng_dwithin(bng_ref, d = 1000)
```

## Grids

`osbng` provides additional several convenience functions to generate BNG grid
square data within specified bounds as a spatial data frame. These functions
require the `sf` package to be available. Grid square data covering the BNG
index system bounds is provided at 100km, 50km, 10km, 5km and 1km resolutions.
The functions are all named `bng_grid_<resolution>()`. Optionally, these
functions can take a bound box argument to restrict the grids to specific
bounds. Resolutions finer than 1km are not supported in these functions to avoid
accidental memory intensive processing.

```{r}
# Create an sf data frame of 100km resolution BNGReference object features
gdf <- bng_grid_100km()

gdf
```

```{r}
# Create an sf data frame of 1km resolution BNGReference object features within a bunding box

# Custom bounding box coordinates
# (xmin, ymin, xmax, ymax)
gdf <- bng_grid_1km(529476, 179654, 532170, 181116)

gdf
```

